Cascaded Approximation of Gaussian Blur

### 2015-05-22 article.tex
Add fifth section

\pgfplotsset{compat=1.9}

@@

>>>>

\pgfplotsset{compat=1.9}
\setlength\intextsep{0pt}

@@

\section{Small blur radius}

The origin of the problem came from the fact that continuous Gaussian function
$\Gamma(\sigma^2;\, x)$ doesn't have direct discrete counterpart for small $\sigma^2$. If we sample
that function at integer arguments as we did in (\ref{gamma}) then we'll have problem with
conservation. Fig.~\ref{norm} shows that sum of all samples diverges from 1 near zero.

\begin{wrapfigure}{r}{0pt}
    \begin{tikzpicture}
        \begin{axis} [ title = $\sum_{k\in Z} \Gamma(\sigma^2;\, k)$,
                width = 6cm, height = 6cm, domain = 0.01:1, samples = 256,
                xmin = 0, xmax = 1, xminorgrids = true, xmajorgrids = true,
                ymin = 0, ymax = 2, yminorgrids = true, ymajorgrids = true,
            ]
            \addplot [ sharp plot, line width = 1pt, red ] gnuplot {
                2 * (0.5 + exp(-1/x/2) + exp(-4/x/2) + exp(-9/x/2) + exp(-16/x/2)) / sqrt(2*pi*x)
            };
        \end{axis}
    \end{tikzpicture}
    \caption{Norm of the sampled Gaussian kernel.}
    \label{norm}
\end{wrapfigure}

One of the possible solutions is to use differential origin of the Gaussian function---diffuse
equation
\begin{equation}
    \frac{\partial\Gamma(\tau;\, x)}{\partial\tau} =
        \frac12 \frac{\partial^2\Gamma(\tau;\, x)}{\partial x^2}.
\end{equation}
We can replace space derivative with corresponding finite difference
\begin{equation}
    \frac{\partial\Gamma^d_n(\tau)}{\partial\tau} =
        \frac12 [\Gamma^d_{n-1}(\tau) - 2\Gamma^d_n(\tau) + \Gamma^d_{n+1}(\tau)]
\end{equation}
and solve that equation. We get
\begin{equation}
    \Gamma^d_n(\tau) = I_n(\tau)e^{-\tau},
\end{equation}
where $I_n(\tau)$ is modified Bessel function
\begin{equation}
    I_n(\tau) = \sum_{k=0}^\infty \frac1{k!(k+n)!}\left(\frac\tau2\right)^{2k+n}.
\end{equation}
Such discrete Gaussian function preserves the main property of its continuous counterpart:
combination of two filters with dispersion $\tau_1$ and $\tau_2$ is filter with dispersion
$\tau = \tau_1 + \tau_2$.

\begin{wrapfigure}{r}{0pt}
    \begin{tikzpicture}
        \begin{axis} [ spectrum, legend cell align = left,
                legend style = { at = {(0.7, 0.5)}, anchor = center } ]
            \addlegendentry{$G(\varphi)$}
            \addplot [ sharp plot, line width = 1pt, red ] gnuplot {
                exp(-0.5*(x*pi)^2)
            };
            \addlegendentry{$\Gamma^d(\sigma^2; \varphi)$}
            \addplot [ sharp plot, line width = 1pt, green ] gnuplot {
                exp(cos(x*pi) - 1)
            };
        \end{axis}
    \end{tikzpicture}
    \caption{Spectrum of different discrete Gaussian approximations for $\sigma^2 = 1$.}
\end{wrapfigure}

Fourier transform of that discrete kernel is
\begin{equation}
    \Gamma^d(\tau;\,\varphi) = e^{\tau(\cos\varphi - 1)}.
\end{equation}
Comparing it ($\tau = \sigma^2$) with (\ref{gauss}) we can see that diffusion-based discrete
Gaussian significantly differs from sampled version even outside of problematic range
$\sigma^2 < 0.5$. Due to that I decided to use sampled version for $\sigma^2 > 0.5$ and extend it
to zero with simple 5-tap filter
\begin{equation}
    G^0_k = \left\{
        \begin{array}{ll}
            1 - \sigma^2 + 6\alpha\sigma^6,& k = 0;\\
            \frac12\sigma^2 - 4\alpha\sigma^6,& k = \pm1;\\
            \alpha\sigma^6,& k = \pm2;\\
            0,& \text{otherwise};
        \end{array}
    \right.
\end{equation}
where $\alpha = 0.085$. Second derivative at zero of Fourier transform of that filter is $\sigma^2$
and that guarantees that combination of such Gaussian filters is Gaussian if the resulting
dispersion is sufficiently large.

\begin{wrapfigure}{r}{0pt}
    \begin{tikzpicture}
        \begin{axis} [ width = 6cm, height = 6cm, domain = 0:1, samples = 256,
                xmin = 0, xmax = 1, xminorgrids = true, xmajorgrids = true,
                ymin = 0, ymax = 1, yminorgrids = true, ymajorgrids = true,
                legend cell align = left ]
            \addlegendentry{$k = 0$}
            \addplot [ sharp plot, line width = 1pt, red] gnuplot {
                x > 0.5 ? exp(-0/x/2) / sqrt(2*pi*x) : 1 - x + 6*0.085*x^3
            };
            \addlegendentry{$k = \pm1$}
            \addplot [ sharp plot, line width = 1pt, green] gnuplot {
                x > 0.5 ? exp(-1/x/2) / sqrt(2*pi*x) : x/2 - 4*0.085*x^3
            };
            \addlegendentry{$k = \pm2$}
            \addplot [ sharp plot, line width = 1pt, blue] gnuplot {
                x > 0.5 ? exp(-4/x/2) / sqrt(2*pi*x) : 0.085*x^3
            };
        \end{axis}
    \end{tikzpicture}
    \caption{Resulting filter coefficients for $0 \le \sigma^2 \le 1$.}
    \label{switch}
\end{wrapfigure}

Terms with $\alpha$ are needed to merge different kernels at $\sigma^2 = 0.5$ without major
discontinuities. Fig.~\ref{switch} shows that junction is quite smooth and even first derivatives
don't change much.

Choosing Gaussian blur representation for small radii concludes the whole algorithm of cascaded
approximation. Resulting algorithm is highly parallel and can be effectively adapted to vector
architectures including modern graphics processors and vector instructions of general-purpose
processors.
